---
title: "Case Study - Cyclistic Data"
author: "Adewale"
output: html_document
date: "2023-05-29"
---

```{r setup, include=FALSE}
## Load the necessary libraries

library(tidyverse) 
library(janitor)
library(skimr)
library(here)
library(dplyr)
```


## Load the csv files into dataframes
```{r csv data collection}
Jun2021 <- read_csv("Raw Data/202106-divvy-tripdata.csv")
```


```{r csv data collection}
Jul2021 <- read_csv("Raw Data/202107-divvy-tripdata.csv")
```


```{r csv data collection}
Aug2021 <- read_csv("Raw Data/202108-divvy-tripdata.csv")
```


```{r csv data collection}
Sep2021 <- read_csv("Raw Data/202109-divvy-tripdata.csv")
```


```{r csv data collection}
Oct2021 <- read_csv("Raw Data/202110-divvy-tripdata.csv")
```


```{r csv data collection}
Nov2021 <- read_csv("Raw Data/202111-divvy-tripdata.csv")
```


```{r csv data collection}
Dec2021 <- read_csv("Raw Data/202112-divvy-tripdata.csv")
```


```{r csv data collection}
Jan2022 <- read_csv("Raw Data/202201-divvy-tripdata.csv")
```


```{r csv data collection}
Feb2022 <- read_csv("Raw Data/202202-divvy-tripdata.csv")
```


```{r csv data collection}
Mar2022 <- read_csv("Raw Data/202203-divvy-tripdata.csv")
```


```{r csv data collection}
Apr2022 <- read_csv("Raw Data/202204-divvy-tripdata.csv")
```


```{r csv data collection}
May2022 <- read_csv("Raw Data/202205-divvy-tripdata.csv")
```


## Combine the loaded data into one dataframe to make analysis easier
```{r}
bike_df <- rbind(Jun2021, Jul2021, Aug2021, Sep2021,
                  Oct2021, Nov2021, Dec2021, Jan2022,
                  Feb2022, Mar2022, Apr2022, May2022)
```
## Create a backup of the combined data
```{r}
bike_df1 <- bike_df
```

## Take a look at the combined data
```{r}
head(bike_df1)
glimpse(bike_df1)
```

## From the above exploration, the started_at and ended_at columns have a character datatype, which is wrong. To proceed, we have to change their datatype to the correct one, datetime.
```{r}
bike_df1[['started_at']] <- as.POSIXct(bike_df1[['started_at']],
format = "%d/%m/%Y %H:%M")

bike_df1[['ended_at']] <- as.POSIXct(bike_df1[['ended_at']],
format = "%d/%m/%Y %H:%M")

glimpse(bike_df1)
head(bike_df1)
```

## Insert new columns which are necessary for the analysis
```{r}
bike_df1$date <- as.Date(bike_df1$started_at) #This will add a date column
bike_df1$year <- format(as.Date(bike_df1$date), "%Y") #This will add a year column
bike_df1$month <- format(as.Date(bike_df1$date), "%B") #This will add a month column
bike_df1$day_of_week[bike_df1$day_of_week=="1"] <- "Sunday"
bike_df1$day_of_week[bike_df1$day_of_week=="2"] <- "Monday"
bike_df1$day_of_week[bike_df1$day_of_week=="3"] <- "Tuesday"
bike_df1$day_of_week[bike_df1$day_of_week=="4"] <- "Wednesday"
bike_df1$day_of_week[bike_df1$day_of_week=="5"] <- "Thursday"
bike_df1$day_of_week[bike_df1$day_of_week=="6"] <- "Friday"
bike_df1$day_of_week[bike_df1$day_of_week=="7"] <- "Saturday"
bike_df1 <- bike_df1 %>%
  rename(weekday = day_of_week)
bike_df1$ride_length <- (as.double(difftime(bike_df1$ended_at, bike_df1$started_at))) /60  #calculate ride length in minutes
head(bike_df1)
```
## Explore the data to see its structure
```{r}
str(bike_df1)
```

```{r}
summary(bike_df1)
```

```{r}
dim(bike_df1)
```


## Data Cleaning

```{r}
#Remove duplicates

bike_df1 <- distinct(bike_df1)

#Some values in the ride_length column are negative, this is impossible. Hence, we have to remove them

bike_df1 <- bike_df1[!bike_df1$ride_length<1,]

#We also need to get rid of ride lengths that are too long, i.e. more than a day or 1440 minutes

bike_df1 <- bike_df1[!bike_df1$ride_length>1440,]

#Some column names do not make sense, we need to change them to give the data they contain meaning

bike_df1 <- bike_df1 %>%
  rename(customer_type = member_casual)

bike_df1 <- bike_df1 %>%
  rename(bike_type = rideable_type)

#We need to focus on the data we will be using for the analysis and filter out unnecessary data

bike_df1 <- bike_df1 %>% select(bike_type, customer_type, started_at, ended_at, start_station_name, end_station_name, date, year, month, weekday, ride_length)

#Order the data based on day of week and by month to make analysis easier
bike_df1$weekday <- ordered(bike_df1$weekday, levels=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
bike_df1$month <- ordered(bike_df1$month, levels=c("June", "July", "August", "September", "October", "November", "December",  "January", "February","March", "April", "May"))

#Remove missing data

bike_df1 <- drop_na(bike_df1)

head(bike_df1)
```

## Analyzing the Data

```{r}
#Let's check membership counts by membership type
table(bike_df1$customer_type)
```
```{r}
#looks at total rides for each customer type in minutes
setNames(aggregate(ride_length ~ customer_type, bike_df1, sum), c("customer_type", "total_ride_length(mins)"))
```

```{r}
#What is the average ride duration?
summary(bike_df1$ride_length)
```

```{r}
bike_df1 %>% group_by(customer_type) %>%
  summarise(min_duration = min(ride_length),
            mean_duration = mean(ride_length),
            median_duration = median(ride_length),
            max_duration = max(ride_length))
```

